Enrichment analysis of differential expression results from AMP-AD (Control vs AD, reprocessed RNASeq)

### 2017-03-01 code/Rmd/MetaAnalysis.Rmd
Minor changes

covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'Control', '4' = 'AD')),
                Study = 'ROSMAP',
                BrainRegion = 'DLPFC') %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 'BM44.SD', 'BM36.ND', 'BM10.ND', 'BM44.ND', 'BM22.ND')) %>%
  dplyr::select(SampleID, BrodmannArea.Dx) %>%
  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Dx, levels = c('ND','SD'), labels = c('ND' = 'Control', 'SD' = 'AD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MAYO',
                Dx = factor(Dx, levels = c('Control','AD')),
                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates = covariates %>%
  rbindlist(use.names = T, fill = T)
```

### Get filtered gene.id and sample.id from synapse
```{r counts}
logcpm.ids = c('ROSMAP' = 'syn8018352',
               'MSBB' = 'syn8073666',
               'MAYO' = 'syn8028565')

logcpm = lapply(logcpm.ids, downloadFile) %>%
  lapply(function(x){
    x = dplyr::select(x, -hgnc_symbol, -ensembl_gene_id)
  }) %>%
  join_all(type = 'full') %>%
  dplyr::select(gene.id, one_of(covariates$SampleID)) 
logcpm[is.na(logcpm)] = min(logcpm[,-(1)], na.rm = T)
```

### Get raw counts from synapse
```{r raw.counts, eval = F}
raw.counts.ids = c('MAYO_CBE' = 'syn7513807',
                   'MSBB' = 'syn7994853',
                   'MAYO_TCX' = 'syn7987765',
                   'ROSMAP' = 'syn7750270')

lcpm = lapply(raw.counts.ids, function(id, lcpm){
  expr = read.table(synGet(id)@filePath, sep = '\t', check.names = F) 
  lib.size = calcNormFactors(expr, method = 'TMM')
  expr = voom(expr, lib.size = lib.size, design = NULL, plot = F, normalize.method = 'quantile')$E
  expr = expr[row.names(expr) %in% lcpm$gene.id, ] %>%
    rownameToFirstColumn('gene.id')
}, logcpm) %>%
  join_all(type = 'full')
```

### Get residual expression from synapse
```{r resid.expr}
# Get SVA adjusted data
resid.expr.id = c('ROSMAP' = 'syn6129372',
                  'MSBB' = 'syn7059289',
                  'MAYO' = 'syn6128634')

resid.expr = lapply(resid.expr.id, downloadFile) %>%
  lapply(function(x){
    x = dplyr::select(x, -hgnc_symbol)
    }) %>%
  join_all(type = 'full') %>%
  dplyr::select(Gene.ID, ensembl_gene_id, one_of(covariates$ID)) 
resid.expr[is.na(resid.expr)] = dplyr::select(resid.expr, -Gene.ID, -ensembl_gene_id) %>%
  min(na.rm=T)
```

@@

# Get background gene ids
backgroundGenes = logcpm[,'gene.id', drop = F] %>% 
  dplyr::mutate(ID = gene.id) %>%
  tidyr::separate(ID, c('ensembl_gene_id','position'), sep = '\\.')

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes$ensembl_gene_id,
                       mart = mart)

@@

Ensemble2HGNC = left_join(backgroundGenes,
                          rbindlist(list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
                            unique) %>%
  unique
```

### Get differential expression results from synapse
```{r diffexp, fig.height=16, fig.width=10}
diff.exp.ids = c(ROSMAP = 'syn8018358',
                 MSBB = 'syn8073678',
                 MAYO = 'syn8028572')
diff.exp = lapply(diff.exp.ids, downloadFile)

# Filter case-control differences alone
diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')

writeLines('Total number of genes differentially expressed between Control-AD at an FDR of 0.05 and foldchange of 1.2 are')
lapply(diff.exp, function(x){
  x = filter(x, adj.P.Val <= 0.05, abs(logFC) >= log2(1.2)) 
  x$Direction = 'NONE'
  x$Direction[x$logFC > 0] = 'UP'
  x$Direction[x$logFC < 0] = 'DOWN'
  x = x %>% group_by(Study, Region, Direction) %>%
    summarise(count = length(unique(gene.id)))
}) %>%
  rbindlist() %>%
  spread(Direction, count) %>%
  kable

tmp = diff.exp %>% 
  rbindlist(use.names = T, fill = T) 
tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC > log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC < -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

pl = list()
for (study in c('ROSMAP','MSBB','MAYO')){
  p = ggplot(tmp %>% filter(Study == study) %>% droplevels, aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme_bw() + theme(legend.position = 'top')
  pl[[study]] = p
}
multiplot(plotlist = pl, cols = 1)

# Find genes that are differentially expressed in all brain regions
diff.exp.gl = dlply(tmp, .(Region), .fun = function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
    dplyr::select(gene.id) %>%
    unique %>% unlist
})

tmp1 = filter(tmp, gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
  dplyr::select(-ensembl_gene_id, -hgnc_symbol) %>%
  left_join(Ensemble2HGNC %>% dplyr::select(-position))
tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'

tmp.eff.sz = group_by(tmp1, gene.id) %>% summarise(es = sum(logFC, na.rm = T)) %>% top_n(6, abs(es))

writeLines('Top six differentially expressed genes are')
pl = list()
p = ggplot(tmp1 %>% filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top')
pl[[1]] = p

@@

multiplot(plotlist = pl, cols = 2)

# Genes differentially expressed in all brain regions
tmp1 = dlply(tmp, .(Region), .fun = function(x){
  dplyr::filter(x, abs(logFC) >= log2(1) & adj.P.Val <= 0.05) %>%
    dplyr::select(gene.id) %>%
    unique %>% unlist
}) %>%
  Reduce(intersect, .)
writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 in all brain regions')
kable(data.frame(ngenes = length(tmp1),
                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))

tmp1 = dlply(tmp, .(Region), .fun = function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05) %>%
    dplyr::select(gene.id) %>%
    unique %>% unlist
}) %>%
  Reduce(intersect, .)
writeLines('Genes that are differentially expressed between Control-AD at FDR 0.05 and foldchange of 1.2 in all brain regions')
kable(data.frame(ngenes = length(tmp1),
                 hgnc_symbol = paste(unique(Ensemble2HGNC$hgnc_symbol[Ensemble2HGNC$gene.id %in% tmp1]), collapse = ', ')))
```

>>>>

covariates$ROSMAP = dplyr::filter(covariates$ROSMAP, cogdx %in% c(1,4)) %>%
  dplyr::mutate(Dx = factor(cogdx, levels = c(1,4), labels = c('1' = 'NCI', '4' = 'AD')),
                Study = 'ROSMAP',
                BrainRegion = 'DLPFC') %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MSBB = dplyr::filter(covariates$MSBB, BrodmannArea.Dx %in% c('BM10.SD', 'BM22.SD', 'BM36.SD', 
                                                                        'BM44.SD', 'BM36.ND', 'BM10.ND', 
                                                                        'BM44.ND', 'BM22.ND')) %>%
  dplyr::select(SampleID, BrodmannArea.Dx) %>%
  tidyr::separate(BrodmannArea.Dx, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MSSM',
                Dx = factor(Dx, levels = c('ND','SD')),
                BrainRegion = factor(BrainRegion, labels = c('FP', 'STG', 'PHG', 'IFG'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)

covariates$MAYO = dplyr::filter(covariates$MAYO, BrainRegion.Diagnosis %in% c('TCX.AD', 'TCX.Control', 'CER.AD', 'CER.Control')) %>%
  dplyr::select(SampleID, BrainRegion.Diagnosis) %>%
  tidyr::separate(BrainRegion.Diagnosis, c('BrainRegion', 'Dx')) %>%
  dplyr::mutate(Study = 'MAYO',
                Dx = factor(Dx, levels = c('Control','AD')),
                BrainRegion = factor(BrainRegion, levels = c('CER', 'TCX'))) %>%
  dplyr::select(SampleID, Study, BrainRegion, Dx)
```

### Get differential expression data from synapse
```{r diffexp}
diff.exp.ids = c(ROSMAP = 'syn8018358',
                 MSBB = 'syn8073678',
                 MAYO = 'syn8028572')
diff.exp = lapply(diff.exp.ids, downloadFile)

diff.exp$ROSMAP = filter(diff.exp$ROSMAP, Comparison == 'cogdx1-cogdx4')
diff.exp$MSBB = filter(diff.exp$MSBB, Comparison == 'ND-SD')
diff.exp$MAYO = filter(diff.exp$MAYO, Comparison == 'Control-AD')

tmp = diff.exp %>% 
  rbindlist(use.names = T, fill = T) 
tmp$Direction = 'NONE'
tmp$Direction[tmp$logFC >= log2(1.2) & tmp$adj.P.Val <= 0.05] = 'UP'
tmp$Direction[tmp$logFC <= -log2(1.2) & tmp$adj.P.Val <= 0.05] = 'DOWN'

writeLines('Number of differentially expressed genes at an FDR of 0.05 and fold change of 1.2 are')
tmp %>%
  group_by(Study, Region, Direction, Comparison) %>%
  summarise(count = n()) %>%
  spread(Direction, count) %>%
  kable

pl = list()
for(region in c('ROSMAP', 'MSBB', 'MAYO')){
  p = ggplot(tmp %>% filter(Study == region), aes(x = logFC, y = -log10(adj.P.Val), color = Direction)) + geom_point()
  p = p + scale_color_manual(values = c('green','grey','red'))
  p = p + facet_grid(Study~.+Region) + theme(legend.position = 'top')
  pl[[region]] = p
}
multiplot(plotlist = pl, cols = 1)
```

### Get cell type fraction from synapse
```{r celltype.frac}
celltype.fraction.ids = c(ROSMAP = 'syn8124520',
                          MSBB = 'syn8146376',
                          MAYO = 'syn8135250')
celltype.fraction = lapply(celltype.fraction.ids, downloadFile)

cell.frac = mapply(function(cov, frc){
  x = filter(frc, SampleID %in% cov$SampleID) %>%
    dplyr::select(-`P-value`,-RMSE, -Correlation) %>%
    tidyr::gather(celltype, fraction, -SampleID) %>%
    left_join(cov)
}, covariates, celltype.fraction, SIMPLIFY = F) %>%
  rbindlist(use.names = T, fill = T) %>%
  dplyr::mutate(Dx = factor(Dx, levels = c('Control', 'NCI', 'ND', 'SD', 'AD')))

pl = list()

p = ggplot(cell.frac %>% filter(celltype %in% c('neurons', 'microglia', 'astrocytes')), 
           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
p = p + facet_grid(celltype~., scales = 'free_y')
p = p + xlab('Brain Region') + ylab('Fraction')
p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
pl[[1]] = p

p = ggplot(cell.frac %>% filter(celltype %in% c('endothelial', 'oligodendrocytes', 'OPC')), 
           aes(x = BrainRegion, y = fraction, color = Dx)) + geom_boxplot()
p = p + facet_grid(celltype~., scales = 'free_y')
p = p + xlab('Brain Region') + ylab('Fraction')
p = p + theme(axis.text.x = element_text(hjust = 1, angle = 90), legend.position = 'top')
pl[[2]] = p

multiplot(plotlist = pl, cols = 2)
```

@@

# Get background gene ids
backgroundGenes = sapply(diff.exp, function(x) x$ensembl_gene_id) %>%
  Reduce(c,.)

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = backgroundGenes,
                       mart = mart)

@@

Ensemble2HGNC = rbindlist( list(Ensemble2HGNC, HGNC2Ensembl), use.names = T, fill = T) %>%
  unique
```

### Consistently differentially expressed genes (in all 7 brain regions)
```{r diffexp.all, fig.height=16, fig.width=10}
# Find genes that are differentially expressed in each brain region
diff.exp.gl = lapply(diff.exp, function(x){
  dplyr::filter(x, abs(logFC) >= log2(1.2) & adj.P.Val <= 0.05 & sign(CI.L) == sign(CI.R)) %>%
    dplyr::select(Region, gene.id) %>%
    unique
}) %>%
  rbindlist(use.names = T, fill = T) %>%
  dlply(.(Region), .fun = function(x){x$gene.id})

tmp1 = rbindlist(diff.exp, use.names = T, fill = T) %>%
  dplyr::filter(gene.id %in% Reduce(intersect, diff.exp.gl)) %>%
  dplyr::select(-hgnc_symbol) %>%
  left_join(Ensemble2HGNC)
tmp1$hgnc_symbol[tmp1$ensembl_gene_id == 'ENSG00000275830'] = 'LOC101927646'

tmp = group_by(tmp1, gene.id) %>%
  summarise(count = sum(logFC/abs(logFC))) %>%
  filter(abs(count) == 7) %>%
  dplyr::mutate(GeneID = gene.id) %>%
  tidyr::separate(gene.id, c('ensembl_gene_id','position'), sep = '\\.') %>%
  left_join(Ensemble2HGNC)
writeLines(paste('Differentially expressed genes in all 7 brain regions at an FDR of 0.05 and fold change of 1.2 are',
                 length(unique(tmp$GeneID))))
kable(data.frame(ngenes = length(unique(tmp$GeneID)),
                 hgnc_symbol = tmp$hgnc_symbol %>% unique %>% paste(collapse = ',')))

writeLines('Top six differentially expressed genes are')
tmp.eff.sz = group_by(tmp1, gene.id) %>% 
  summarise(es = sum(logFC, na.rm = T)) %>% 
  top_n(6, abs(es))

pl = list()
p = ggplot(tmp1 %>% 
             filter(gene.id %in% tmp.eff.sz$gene.id[tmp.eff.sz$es>0]), 
           aes(x = Region, y = logFC, fill = Study)) + geom_bar(stat = 'identity')
p = p + facet_grid(hgnc_symbol~.) + geom_errorbar(aes(ymin=CI.L, ymax=CI.R), width = 0.2)
p = p + theme(legend.position = 'top')
pl[[1]] = p

@@

multiplot(plotlist = pl, cols = 2)
```

### 2017-03-01 code/Rmd/enrichmentAnalysis.Rmd
Minor changes

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}
## It is assumed your working directory is where this file is

@@

library(Vennerable)
library(colorfulVennPlot)

@@

# Filter results
tmp1 = filter(enrichResults, fdr <= 0.05)
```

@@

# Store enrichment analysis in synapse
write.table(tmp1, file = 'enrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrichmentResults.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(ALL_USED_IDs),
               executed = thisFile,
               activityName = 'Pathway enrichment analysis')
stopCluster(cl)
```

>>>>

```{r libs, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE, cache=FALSE}
## It is assumed your working directory is where this file is

@@

library(Vennerable)

@@

# Filter results
tmp1 = filter(enrichResults, fdr <= 0.05)

# Plot results
tmp = enrichResults %>%
  dplyr::filter(Name %in% tmp1$Name) %>%
  dplyr::select(Region, Name, fdr) %>% 
  tidyr::spread(Region, fdr)
rownames(tmp) = tmp$Name
tmp = data.matrix(tmp[,-(1)])
tmp[is.na(tmp)] = 1
tmp = tmp[setdiff(1:dim(tmp)[1], grep('Mus', rownames(tmp))),]

Heatmap(tmp, name = 'fdr', show_row_names = F)

# In any 3
h = Heatmap(tmp[rowSums(tmp <= 0.05, na.rm = T) >= 3,], name = 'fdr',
        col = colorRamp2(c(0,1), c('red', 'white')),
        show_row_dend = F, row_names_side = 'right',
        row_names_gp = gpar(fontsize = 8))
draw(h, heatmap_legend_side = 'left')

# Overlap with AMP-AD targets
targets = downloadFile('syn7555804') %>%
  dplyr::select(GENE_SYMBOL) %>%
  unlist() %>% unique()

plyr::ldply(diff.exp.gl, .fun = fisherEnrichment, targets, backgroundGenes, .id = 'Name', .parallel = T) %>%
  kable

```

@@

# Store enrichment analysis in synapse
write.table(tmp1, file = 'enrichmentResults.tsv', sep = '\t', row.names=F, quote=F)
OBJ = File('enrichmentResults.tsv', 
           name = 'Pathway Enrichment Results (AD and Control)', 
           parentId = CODE$properties$id)
OBJ = synStore(OBJ,
               used = as.character(ALL_USED_IDs),
               executed = thisFile,
               activityName = 'Pathway enrichment analysis')
# stopCluster(cl)
```
